Hockey's Most Controversial Statistic: An Analysis of the Effectiveness of the Plus-Minus Statistic

### 2024-12-05 manuscript/statspaper.tex
changed the format of the tables to fit

\begin{abstract}
The goal of this paper is to evaluate the plus minus statistic as a metric for determining the effectiveness of a player.
This involves analysing the relationship between a players plus-minus rating and other offensive and defensive ratings,
the effect of a players team performance on their plus-minus, and comparing plus-minus to other similar statistics. Based
on the analysis performed, the plus minus does have some value as a predictor of offensive and defensive metrics. However, other
advanced metrics like Corsi and Fenwick perform much better. The plus-minus is also heavily influenced by a player's team, making it 
less valuable as an indivual evaluator of player effectiveness.
\end{abstract}

@@

While the plus-minus statistic is a great idea in theory, it does not come without its weaknesses. The biggest drawback to this
statistic is its issues with independence. Many ice hockey players tend to play on "lines", meaning the same three 
forwards tend to play together and the same two defensemen usually play together. As a result, the performance of one player is highly dependent 
on the performance of their linemates. In "A Regression-based Adjusted Plus-Minus Statistic for NHL Players" \cite{Macdonald_2011}, Brian
MacDonald provides a perfect example of this with the Henrik and Daniel Sedin, Swedish twins who played together for the Vancouver Canucks. 
 \cite[Daniel spent 92\% of his playing time with Henrik, the highest percentage of any other player combination where both players have
played over 700 minutes. Because of this high collinearity between the twins, it is difficult to separate the individual effect that each 
player has on the net goals scored on the ice.]{Macdonald_2012} Many critics of the plus-minus argue that plus-minus is more of a team 
statistic, since it is heavily influenced by team dynamics rather than individual contributions. Furthermore, numerous confounding variables, 
such as the quality of the opponent and situational factors, make the calculation less reliable. As a result of these criticisms, alternatives 
to the plus-minus statistic have been developed, such as Corsi or Fenwick, which is calculated using shot attempts rather than goals.

@@

\subsection{Correlation Analysis}
The main issue with the data is multicollinearity, where independent variables are highly correlated. A correlation matrix was created 
with some of the more basic nhl statistics to highlight this issue, shown in Table \ref{tab:correlation_matrix}.

\begin{table}[h!]
  \centering
  \caption{Correlation Matrix of Variables}
  \input{correlation_matrix.tex}
  \label{tab:correlation_matrix}
\end{table}

@@

\begin{table}[tbp]
  \caption{Correlation Coefficients with Plus-Minus}
  \label{tab:rv}
\centering
\begin{tabular}{rr}
  \toprule
Variable & Correlation Coefficient \\ 
  \midrule
  GP & 0.232 \\ 
  G & 0.237 \\ 
  A & 0.343 \\ 
  P & 0.315 \\ 
  +/- & 1.000 \\ 
  P/GP & 0.278 \\ 
  EVG & 0.239 \\ 
  EVP & 0.336 \\ 
  PPG & 0.187 \\ 
  PPP & 0.227 \\ 
  SHG & 0.177 \\ 
  SHP & 0.283 \\ 
  OTG & 0.162 \\ 
  GWG & 0.306 \\ 
  S & 0.256 \\ 
  S\% & 0.048 \\ 
  TOI/GP & 0.253 \\ 
  TOI & 0.292 \\ 
  CF & 0.369 \\ 
  CA & 0.211 \\ 
  CF\% & 0.534 \\ 
  FF & 0.374 \\ 
  FA & 0.213 \\ 
  FF\% & 0.554 \\ 
  SF & 0.376 \\ 
  SA & 0.212 \\ 
  SF\% & 0.569 \\ 
  GF & 0.424 \\ 
  GA & 0.083 \\ 
  GF\% & 0.705 \\ 
  xGF & 0.387 \\ 
  xGA & 0.196 \\ 
  xGF\% & 0.586 \\ 
  SCF & 0.387 \\ 
  SCA & 0.192 \\ 
  SCF\% & 0.585 \\ 
  HDCF & 0.399 \\ 
  HDCA & 0.196 \\ 
  HDCF\% & 0.551 \\ 
  HDGF & 0.442 \\ 
  HDGA & 0.093 \\ 
  HDGF\% & 0.614 \\ 
  On-Ice SH\% & 0.324 \\ 
  On-Ice SV\% & 0.419 \\ 
  PDO & 0.546 \\ 
  Off. Zone Starts & 0.310 \\ 
  Neu. Zone Starts & 0.295 \\ 
  Def. Zone Starts & 0.237 \\ 
  On The Fly Starts & 0.273 \\ 
  Off. Zone Start \% & 0.111 \\ 
  Off. Zone Faceoffs & 0.354 \\ 
  Neu. Zone Faceoffs & 0.315 \\ 
  Def. Zone Faceoffs & 0.263 \\ 
  Off. Zone Faceoff \% & 0.183 \\ 
  \bottomrule
\end{tabular}
\end{table}

>>>>

\begin{abstract}

The goal of this paper is to evaluate the plus minus statistic as a metric for determining the effectiveness of a player.
This involves analysing the relationship between a players plus-minus rating and other offensive and defensive ratings,
the effect of a players team performance on their plus-minus, and comparing plus-minus to other similar statistics. Based
on the analysis performed, the plus minus does have some value as a predictor of offensive and defensive metrics. However, other
advanced metrics like Corsi and Fenwick perform much better. The plus-minus is also heavily influenced by a player's team, making it 
less valuable as an indivual evaluator of player effectiveness.

\end{abstract}

@@

While the plus-minus statistic is a great idea in theory, it does not come without its weaknesses. The biggest drawback to this
statistic is its issues with independence. Many ice hockey players tend to play on "lines", meaning the same three 
forwards tend to play together and the same two defensemen usually play together. As a result, the performance of one player is highly dependent 
on the performance of their linemates. In "A Regression-based Adjusted Plus-Minus Statistic for NHL Players" \cite{Macdonald_2011}, Brian
MacDonald provides a perfect example of this with the Henrik and Daniel Sedin, Swedish twins who played together for the Vancouver Canucks. 
"Daniel spent 92\% of his playing time with Henrik, the highest percentage of any other player combination where both players have
played over 700 minutes. Because of this high collinearity between the twins, it is difficult to separate the individual effect that each 
player has on the net goals scored on the ice."\cite{Macdonald_2012} Many critics of the plus-minus argue that plus-minus is more of a team 
statistic, since it is heavily influenced by team dynamics rather than individual contributions. Furthermore, numerous confounding variables, 
such as the quality of the opponent and situational factors, make the calculation less reliable. As a result of these criticisms, alternatives 
to the plus-minus statistic have been developed, such as Corsi or Fenwick, which is calculated using shot attempts rather than goals. Brian 
MacDonald introduces an adjusted plus-minus statistic that is based on weighted least squares regression\cite{Macdonald_2012}.

@@

\subsection{Correlation Analysis}
The main issue with the data is multicollinearity, where independent variables are highly correlated. A correlation matrix was created 
with some of the more basic nhl statistics to highlight this issue, shown in Table \ref{tab:correlation_matrix2}.

\begin{table}[h!]
  \centering
  \scriptsize
  \caption{Correlation Matrix of Variables}
  \input{correlation_matrix2.tex}
  \label{tab:correlation_matrix2}
\end{table}

@@

\begin{table}[tbp]
  \caption{Correlation Coefficients with Plus-Minus}
  \label{tab:rv}
\centering
\small
\begin{tabular}{rr}
  \toprule
Variable & Correlation Coefficient \\ 
  \midrule
  GP & 0.232 \\ 
  G & 0.237 \\ 
  A & 0.343 \\ 
  P & 0.315 \\ 
  +/- & 1.000 \\ 
  P/GP & 0.278 \\ 
  EVG & 0.239 \\ 
  EVP & 0.336 \\ 
  PPG & 0.187 \\ 
  PPP & 0.227 \\ 
  SHG & 0.177 \\ 
  SHP & 0.283 \\ 
  OTG & 0.162 \\ 
  GWG & 0.306 \\ 
  S & 0.256 \\ 
  S\% & 0.048 \\ 
  TOI/GP & 0.253 \\ 
  TOI & 0.292 \\ 
  CF & 0.369 \\ 
  CA & 0.211 \\ 
  CF\% & 0.534 \\ 
  FF & 0.374 \\ 
  FA & 0.213 \\ 
  FF\% & 0.554 \\ 
  SF & 0.376 \\ 
  SA & 0.212 \\ 
  SF\% & 0.569 \\ 
  GF & 0.424 \\ 
  GA & 0.083 \\ 
  GF\% & 0.705 \\ 
  xGF & 0.387 \\ 
  xGA & 0.196 \\ 
  xGF\% & 0.586 \\ 
  SCF & 0.387 \\ 
  SCA & 0.192 \\ 
  SCF\% & 0.585 \\ 
  HDCF & 0.399 \\ 
  HDCA & 0.196 \\ 
  HDCF\% & 0.551 \\ 
  HDGF & 0.442 \\ 
  HDGA & 0.093 \\ 
  HDGF\% & 0.614 \\ 
  On-Ice SH\% & 0.324 \\ 
  On-Ice SV\% & 0.419 \\ 
  PDO & 0.546 \\ 
  \bottomrule
\end{tabular}
\end{table}
